---
title: "Getting started with tensorsl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tensorsl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tensorsl)
```

## What is TSL

TSL (Tensor Separation Learning) is a glass-box regression model. A fitted model
is a sum of *stages*; each stage is the ordered difference of two non-negative
rank-1 products of univariate functions. Concretely, a stage predicts
$f = \lambda_+ \prod_j b_j e^{d_j} - \lambda_- \prod_j b_j e^{-d_j}$, where the
per-feature backbone $b_j \ge 0$ is a shared magnitude (a gate) and the tilt
$d_j \in \mathbb{R}$ is the signed direction. Because every piece is univariate
and separable, you can read off what the model learned feature by feature.

## Fit and predict

We build a synthetic, mostly-separable regression problem on a `400 x 3` matrix
with named columns. The true response is roughly additive in three features, so
the glass-box story is easy to check.

```{r data}
set.seed(1)
n <- 400
x <- matrix(runif(n * 3, -2, 2), ncol = 3,
            dimnames = list(NULL, c("a", "b", "c")))
y <- 2 * x[, "a"] - x[, "b"] + 0.5 * x[, "c"] + rnorm(n, sd = 0.1)
```

Split into train and test sets, fit a small boosted model, and predict.

```{r fit}
train <- 1:300
test  <- 301:400

model <- tsl(x[train, ], y[train],
             epochs = 8L, n_trees = 6L, seed = 1L, verbosity = 0L)
print(model)

pred <- predict(model, x[test, ])
rmse <- sqrt(mean((y[test] - pred)^2))
rmse
```

## Inspecting the glass box

`tsl_components()` returns the fitted structure in two-tensor form, with one
entry per boosting stage.

```{r components}
comp <- tsl_components(model)
length(comp)            # one entry per stage
str(comp[[1]], max.level = 1)
```

Per stage, `scaling_plus` and `scaling_minus` are the OLS coefficients on the
$+$ and $-$ branches -- the only place scaling is applied. `combined_grid_tensor`
is the aggregated representative component for the stage, and `grid_tensors` is
the per-tree bag it was aggregated from.

Each grid tensor stores, per feature, the interval `splits`, the non-negative
gate magnitude `backbone_values` ($b \ge 0$), the signed direction
`tilt_values` ($d$), and the `observation_counts` per interval, plus the branch
scalars `lambda_plus` and `lambda_minus`.

```{r grid-tensor}
str(comp[[1]]$grid_tensors[[1]], max.level = 1)
```

## Marginal effect plot

To see what the model learned about a single feature, sweep that feature across
a grid while holding the others at their medians, then predict along the sweep.
Here we recover the (roughly linear) effect of feature `a`.

```{r marginal, fig.width = 6, fig.height = 4}
grid_a <- seq(-2, 2, length.out = 100)
medians <- apply(x[train, ], 2, median)

newx <- matrix(rep(medians, each = length(grid_a)), ncol = 3,
               dimnames = list(NULL, c("a", "b", "c")))
newx[, "a"] <- grid_a

plot(grid_a, predict(model, newx), type = "l",
     xlab = "a", ylab = "predicted f",
     main = "Marginal effect of feature a")
```

The response rises monotonically with `a`, matching the true `2 * a` term.

## Reproducibility and Python parity

The hyperparameters mirror the Python `TSLRegressor`, so a fit with the same
data and `seed` reproduces the Python results. Within R, two fits with the same
seed are identical down to the fitted components.

```{r reproducibility}
m1 <- tsl(x[train, ], y[train], epochs = 8L, n_trees = 6L, seed = 1L, verbosity = 0L)
m2 <- tsl(x[train, ], y[train], epochs = 8L, n_trees = 6L, seed = 1L, verbosity = 0L)
identical(tsl_components(m1), tsl_components(m2))
```
